---
title: "Creating Pre-attentive attributes"
author: "Tim Essam"
format: html
editor: visual
---

## Pre-attentive attributes in ggplot2 and R

In the code below, we review a few of the pre-attentive attributes covered in our lecture.

To learn more about ggplot2's asthetics see the `vignette("ggplot2-specs")` document.

Load the PEPFAR data into your work space.

```{r prep}
#| echo: false

# Load the libraries needed
library(glitr)
library(glamr)
library(tidyverse)

# Load data, munge down to smaller dataset
#    df_msd <- return_latest("Data", "PSNU_IM") %>%
#       read_psd() %>% 
#       filter(indicator == "HTS_TST_POS", 
#              standardizeddisaggregate == "Total Numerator")
#   
# write_rds(df_msd, "Data/df_msd.rds")

# Read in MSD data - simplified to only testing data
dir()
df_msd <- read_rds("Data_public/df_msd.rds")
  
# Subset the MSD for examples
df_tst_psnu <- df_msd %>%
  filter(indicator == "HTS_TST_POS") %>%
  summarize(
    cumulative = sum(cumulative, na.rm = T),
    .by = c("fiscal_year", "indicator", "psnu")
  ) %>%
  arrange(psnu) %>%
  mutate(fy = as.character(fiscal_year)) %>% 
  slice(1:9)

write_rds(df_tst_psnu, "Data_public/df_tst_psnu.rds")

# What does this dataset look like?
df_tst_psnu
glimpse(df_tst_psnu)
  
```

### Color

```{r color}
# Set height and width
  h <- 3.25
  w <- 5.75

# geom_col
df_tst_psnu %>%
  ggplot(aes(
    x = fy,
    y = cumulative,
    color = psnu
  )) +
  geom_point(size = 4)
si_save("Images/ggplot_aes_color.png", height = h, width = w)
```

### Size

```{r size}
# Size
df_tst_psnu %>%
  ggplot(aes(
    x = fy,
    y = cumulative,
    size = cumulative
  )) +
  geom_point()
si_save("Images/ggplot_aes_size.png", height = h, width = w)
```

### Shape

```{r shape}
df_tst_psnu %>%
  ggplot(aes(
    x = fy,
    y = cumulative,
    shape = psnu
  )) +
  geom_point(size = 4, position = position_jitter(width = 0.15))
si_save("Images/ggplot_aes_shape", height = h, width = w)
```

### Saturation

```{r saturation}
df_tst_psnu %>%
  ggplot(aes(
    x = fy,
    y = cumulative,
    alpha = cumulative
  )) +
  geom_point(size = 5)
si_save("Images/ggplot_aes_transparency", height = h, width = w)
```

### Enclosure

```{r enclosure}
df_tst_psnu %>%
  ggplot(aes(
    x = fy,
    y = cumulative,
  )) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 1200, ymax = 1600, alpha = .15) +
  geom_point(size = 5) 
si_save("Images/ggplot_aes_enclosure", height = h, width = w)
```

### Putting them all together

```{r combo}
# Combining color, shape and size
df_tst_psnu %>%
  ggplot(aes(
    x = fy,
    y = cumulative,
    color = psnu,
    size = cumulative,
    shape = psnu
  )) +
  geom_point()
si_save("Images/ggplot_aes_combining", height = h, width = w)
```

### Applying glitr colors

```{r genoa}
# Combine it with glitr colors and themes
df_tst_psnu %>%
  ggplot(aes(x = fy, y = cumulative, color = psnu)) +
  geom_point(size = 4, position = position_jitter(width = 0.1, seed = 42)) + 
  scale_color_si(palette = "genoa", discrete = T) +
  labs(title = "ALBUQUERQUE LAGS BEHIND IN TESTING",
       subtitle = "2060 results through Q2",
       caption = "Faux MSD 2023")

si_save("Images/ggplot_glitr_colors_genoa", height = h, width = w)

# si_palettes$genoas %>% scales::show_col(ncol = 5)
# si_palettes$denim %>% scales::show_col(ncol = 5)
# si_palettes$denims %>% scales::show_col(ncol = 5)
```

```{r denim}
df_tst_psnu %>%
  ggplot(aes(x = fy, y = cumulative, color = psnu)) +
  geom_point(size = 4, position = position_jitter(width = 0.1,
                                                  seed = 42)) + 
  scale_color_si(palette = "denim", discrete = T) +
  labs(title = "ALBUQUERQUE LAGS BEHIND IN TESTING",
       subtitle = "2060 results through Q2",
       caption = "Faux MSD 2023")
si_save("Images/ggplot_glitr_colors_denim", height = h, width = w)
```

```{r denims}
df_tst_psnu %>%
  ggplot(aes(x = fy, y = cumulative, color = cumulative)) +
  geom_point(size = 4, position = position_jitter(width = 0.1, seed = 42)) + 
  scale_color_si(palette = "denims") +
  labs(title = "ALBUQUERQUE LAGS BEHIND IN TESTING",
       subtitle = "2060 results through Q2",
       caption = "Faux MSD 2023")
si_save("Images/ggplot_glitr_colors_denims", height = h, width = w)
```

### Reducing clutter with `si_style()`

```{r si_style}
# Modify the style
df_tst_psnu %>%
  ggplot(aes(x = fy, y = cumulative, color = psnu)) +
  geom_point(size = 4, position = position_jitter(width = 0.1, seed = 42)) + 
  scale_color_si(palette = "genoa", discrete = T) +
  labs(title = "ALBUQUERQUE LAGS BEHIND IN TESTING",
       subtitle = "2060 results through Q2",
       caption = "Faux MSD 2023") +
  si_style()
si_save("Images/ggplot_glitr_colors_si_style", height = h, width = w)
```

```{r si_style_ygrid}
df_tst_psnu %>%
  ggplot(aes(x = fy, y = cumulative, color = psnu)) +
  geom_point(size = 4, position = position_jitter(width = 0.1, seed = 42)) + 
  scale_color_si(palette = "genoa", discrete = T) +
  labs(title = "ALBUQUERQUE LAGS BEHIND IN TESTING",
       subtitle = "2060 results through Q2",
       caption = "Faux MSD 2023") +
  si_style_ygrid() 
si_save("Images/ggplot_glitr_colors_si_style_ygrid", height = h, width = w)
```

```{r si_style_xline}
df_tst_psnu %>%
  ggplot(aes(x = fy, y = cumulative, color = psnu)) +
  geom_point(size = 4, position = position_jitter(width = 0.1, seed = 42)) + 
  si_style_ygrid() +
  scale_color_si(palette = "genoa", discrete = T) +
  labs(title = "ALBUQUERQUE LAGS BEHIND IN TESTING",
       subtitle = "2060 results through Q2",
       caption = "Faux MSD 2023") +
  si_style_xline()
si_save("Images/ggplot_glitr_colors_si_style_xline", height = h, width = w)
```
